---
title: "Getting started with decideR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with decideR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); macOS arm64, R 4.5.1, one
     core. No sampler runs in this vignette; every decision is a closed-form
     expected-utility scan over a candidate grid. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
set.seed(1L)
```

```{r library}
library(decideR)
```

## Why

An agronomist closing out a nitrogen trial holds posterior draws of grain
yield at a handful of candidate rates and one concrete question to answer for
the grower: *what rate should go on, and how much should I trust that
number?* `decideR` answers both halves of that question with one call: it
turns a posterior into the profit-maximising rate over the candidates on
offer, and it declines to recommend a rate at all when the yield evidence
behind the posterior has not been checked against its source. This vignette
walks that nitrogen decision from posterior draws to a marked-up candidate
curve, shows the refusal for real on the identical draws, and then carries
the same engine into two further settings -- a trading position sized under
a capital-preservation constraint, and a grain-quality payoff that steps
rather than rises smoothly -- because the underlying question, act under
uncertainty or decline to, is the same question in every one of them.

## What

`decide()` is the shared keystone every decision in this vignette reduces
to. Given posterior `draws` of a latent state, a `utility(action, theta)`
function, and a set of candidate actions, it evaluates the expected utility
of each feasible candidate by averaging over the draws and returns a
`decision`: an S7 object carrying the chosen `action`, the
`expected_utility` that justified it, the full `candidates` ledger, and two
fields the Independent Oracle Principle requires on every decision --
`grounding` and the abstention record (`abstained`, `abstain_reason`). A
decision abstains, in priority order, when the combined `grounding` is not
grounded, when no candidate is feasible, when the effective sample size
falls below an opt-in `min_ess` floor, or when the leading candidate does
not clearly beat the safe action; a fifth trigger catches a utility that
returns no usable value anywhere on the grid. `grounding_grounded()` and
`grounding_unverified()` are the two tokens a fact can carry, and
`combine_grounding()` folds several into one worst case, so a decision
resting on several inputs is only as grounded as its weakest one. The full
vocabulary of abstention reasons, in the order `decide()` checks them:

```{r abstention-table, echo = FALSE}
abstention_reasons <- data.frame(
  reason = c(
    "input_ungrounded", "no_feasible_action", "low_ess",
    "insufficient_evidence", "degenerate_utility"
  ),
  meaning = c(
    "An input fact carries grounding_unverified() (the IOP firewall).",
    "No candidate satisfies the feasibility constraint.",
    "The effective sample size is below min_ess (opt-in; off by default).",
    "The best action does not clearly beat the safe action.",
    "Every feasible candidate's expected utility is NA or NaN."
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  abstention_reasons, col.names = c("Reason", "Meaning"),
  caption = "The abstention vocabulary decide() reports, checked in order."
)
```

`decide_input_rate()` and `decide_position_size()` are the two reference
decisions this vignette exercises directly: the first chooses an agronomic
input rate by expected profit over a matrix of per-rate yield draws, the
second sizes a trading position by capped expected log-growth (the Kelly
objective) under a hard one-period loss-quantile constraint. Both route
through `decide()`'s shared core, so a nitrogen rate and a trading position
are, structurally, the same computation with a different utility and a
different candidate grid. `decide_from_manifest()` and its agronomic
convenience `decide_rate_from_manifest()` are the manifest-native pipeline
tail: given an `orchestra_manifest` (or any S7 object exposing the same
`outputs`/`metadata`/`summary` properties) emitted upstream, they read the
predictive draws and the producer's own grounding token straight off it, so
a decision at the end of a pipeline never re-derives provenance an upstream
member already established. `grade_band_value()` builds a step value
schedule for a quality-graded payoff -- a grain price that jumps at a
threshold rather than rising continuously -- and `grade_band_utility()`
turns that schedule into a `decide()` utility directly (not exercised below,
where the equivalent shift is written by hand to keep the worked example
self-contained).

## Do

### A nitrogen-rate decision

An upstream model -- a mechanistic-simulator inversion, say -- has produced
posterior draws of grain yield at each of a set of candidate nitrogen rates,
following a diminishing-returns response with posterior noise. The engine
chooses the rate that maximises expected profit,
$\mathrm{E}[p_y\,y(r) - p_r\,r]$, at a grain price $p_y$ and an input price
$p_r$.

```{r nitrogen-toy}
rates <- seq(0, 200, by = 25)
ymax <- rnorm(2000L, mean = 5.5, sd = 0.3)      # posterior of asymptotic yield
yield_draws <- vapply(
  rates,
  function(r) 3 + (ymax - 3) * (1 - exp(-r / 80)) + rnorm(2000L, 0, 0.1),
  numeric(2000L)
)
```

```{r nitrogen-decide}
d_crop <- decide_input_rate(
  yield_draws, rates, price_grain = 300, price_input = 1.2,
  grounding = grounding_grounded()
)
d_crop
```

The full candidate ledger -- every rate evaluated, feasible or not, with its
expected profit -- is available for inspection or a methods appendix rather
than discarded once the maximiser has run.

```{r nitrogen-ledger-kable, echo = FALSE}
knitr::kable(
  d_crop@candidates, digits = 0,
  col.names = c("Rate (kg N/ha)", "Expected profit (AUD/ha)", "Feasible"),
  caption = paste0(
    "Candidate ledger for the nitrogen-rate decision, every rate on ",
    "the grid."
  )
)
```

```{r fig-nitrogen-ledger, echo = requireNamespace("ggplot2", quietly = TRUE), eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Expected profit against candidate nitrogen rate. The curve rises steeply at low rates, where nitrogen is well below the point of diminishing yield response, then flattens as the input cost catches up with the marginal yield gain; the chosen rate (marked) sits at the turning point, not at the top of the grid."}
ledger_df <- d_crop@candidates
ledger_df$point_type <- ifelse(
  ledger_df$action == d_crop@action, "chosen", "evaluated"
)
ggplot2::ggplot(ledger_df, ggplot2::aes(x = action, y = expected_utility)) +
  ggplot2::geom_line(colour = "grey50") +
  ggplot2::geom_point(ggplot2::aes(colour = point_type), size = 3) +
  ggplot2::geom_vline(
    xintercept = d_crop@action, linetype = "dashed", colour = "grey40"
  ) +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.8) +
  ggplot2::labs(
    x = "Nitrogen rate (kg N/ha)", y = "Expected profit (AUD/ha)",
    title = "Candidate ledger: nitrogen rate"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-nitrogen-ledger-skip, echo = FALSE, eval = !requireNamespace("ggplot2", quietly = TRUE), results = "asis"}
cat("ggplot2 is not installed; the candidate-ledger figure is skipped.\n")
```

### The grounding firewall, applied to a nitrogen rate

The identical posterior, prices and candidate grid, but this time the yield
evidence carries `grounding_unverified()` -- not checked against an external
authority, the Independent Oracle Principle default.

```{r nitrogen-ungrounded}
d_crop_ungrounded <- decide_input_rate(
  yield_draws, rates, price_grain = 300, price_input = 1.2,
  grounding = grounding_unverified()
)
d_crop_ungrounded
```

### The same engine, a different domain: sizing a trading position

`decide_position_size()` routes through the identical core: given predictive
draws of the next-period return, it chooses the fraction of capital that
maximises expected log-growth $\mathrm{E}[\log(1 + f\,r)]$ over a grid capped
at `kelly_cap`, subject to a hard one-period loss-quantile constraint.

```{r trading-decide}
returns <- rnorm(5000L, mean = 0.02, sd = 0.03)    # a small positive edge
d_trade <- decide_position_size(
  returns, capital = 1000, max_drawdown = 0.15,
  grounding = grounding_grounded()
)
d_trade
d_trade@metadata$position                          # implied dollar position
```

With no edge, the grid maximiser is naturally flat -- the engine does not
manufacture a position where none is warranted.

```{r trading-flat}
flat <- decide_position_size(
  rnorm(5000L, mean = 0, sd = 0.03), capital = 1000, max_drawdown = 0.15,
  grounding = grounding_grounded()
)
flat@action                                        # 0: flat
```

And the capital firewall: the same return draws that produced a decided
position above, now carrying `grounding_unverified()` -- a price feed not
checked against the exchange -- force flat, regardless of how attractive the
numbers look.

```{r trading-firewall}
d_trade_ungrounded <- decide_position_size(
  returns, capital = 1000, max_drawdown = 0.15,
  grounding = grounding_unverified()
)
d_trade_ungrounded@abstained
d_trade_ungrounded@abstain_reason
```

### The pipeline tail: deciding from an upstream manifest

In an ORCHESTRA pipeline the yield posterior above is not assembled by
hand -- it arrives as an `orchestra_manifest` emitted by an upstream member
(a PESTO ensemble inversion, a kernR causal test, a flexyBayes posterior).
`decide_rate_from_manifest()` reads the predictive draws and the producer's
grounding token straight off the manifest and closes the loop into a
decision. `decideR` does not depend on the contract package -- the manifest
is read through its public properties (`outputs`, `metadata`, `summary`), so
the tail consumes any conforming producer rather than special-casing each
one. In production the manifest is a real `orchestra_manifest` S7 object
from the composition layer; a minimal S7 stand-in carrying the same public
properties is built below to keep this vignette self-contained.

```{r manifest-stand-in}
vignette_manifest <- S7::new_class(
  "vignette_manifest",
  properties = list(
    outputs         = S7::new_property(S7::class_any, default = NULL),
    metadata        = S7::new_property(S7::class_list, default = list()),
    summary         = S7::new_property(S7::class_any, default = NULL),
    run_id          = S7::new_property(S7::class_character, default = ""),
    emitter_package = S7::new_property(S7::class_character, default = "")
  )
)
m_grounded <- vignette_manifest(
  outputs = yield_draws,
  metadata = list(grounding = grounding_grounded()),
  run_id = "pesto-001", emitter_package = "PESTO"
)
m_ungrounded <- vignette_manifest(
  outputs = yield_draws,
  metadata = list(grounding = grounding_unverified()),
  run_id = "pesto-002", emitter_package = "PESTO"
)
```

```{r manifest-tail}
d_tail <- decide_rate_from_manifest(
  m_grounded, rates = rates, price_grain = 300, price_input = 1.2
)
d_tail
d_tail@inputs$emitter_package      # provenance carried from the manifest
```

The grounding rides from the producer through the decision: a manifest whose
payload carries `grounding_unverified()` forces the status quo, the same
firewall applied at the pipeline boundary rather than inside a single call.

```{r manifest-tail-ungrounded}
d_tail_ungrounded <- decide_rate_from_manifest(
  m_ungrounded, rates = rates, price_grain = 300, price_input = 1.2
)
d_tail_ungrounded@abstained
```

For a fully general tail -- any candidate set, any utility --
`decide_from_manifest()` is the function `decide_rate_from_manifest()`
wraps.

### A threshold-economics loss: grade-band payoffs

Many grain payoffs are not smooth in the quality they reward. A delivery is
paid by grade: the price steps up as protein clears a milling threshold and
again at a premium threshold. The value of an outcome depends on which band
it falls in, not on its exact magnitude, so the decision is governed by the
posterior probability of clearing each band, not by the posterior mean.
`grade_band_value()` makes that step payoff a first-class, reusable loss: a
value schedule with breakpoints.

```{r grade-band-schedule}
protein_value <- grade_band_value(
  breaks = c(11.5, 13.0),            # milling and premium protein thresholds
  values = c(300, 360, 420)          # AUD/t: feed, milling, premium
)
protein_value
```

Nitrogen lifts grain protein with diminishing returns; the per-hectare value
folds in a 5.5 t/ha yield, and nitrogen costs 1.3 per unit rate. A posterior
over the base (zero-nitrogen) protein feeds the rate of greatest expected
per-hectare value.

```{r grade-band-decide}
qshift <- function(rate, base_protein) {
  base_protein + 3.0 * (1 - exp(-rate / 50))    # protein response to nitrogen
}
yield <- 5.5
grain_value <- function(rate, base_protein) {
  yield * protein_value(qshift(rate, base_protein)) - 1.3 * rate
}

base_protein <- rnorm(4000L, mean = 10.6, sd = 0.5)
rates_n <- seq(0, 150, by = 15)
d_protein <- decide(
  base_protein, grain_value, candidates = rates_n,
  grounding = grounding_grounded(), safe_action = 0
)
d_protein
```

```{r grade-band-probs}
prot_best <- qshift(d_protein@action, base_protein)
p_milling <- mean(prot_best >= 11.5)
p_premium <- mean(prot_best >= 13.0)
c(`P(>=11.5 milling)` = p_milling, `P(>=13 premium)` = p_premium)
```

```{r grade-band-ledger-kable, echo = FALSE}
knitr::kable(
  d_protein@candidates, digits = 0,
  col.names = c("Rate (kg N/ha)", "Expected value (AUD/ha)", "Feasible"),
  caption = "Candidate ledger for the grade-band nitrogen decision."
)
```

```{r fig-grade-band, echo = requireNamespace("ggplot2", quietly = TRUE), eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "The grade-band step schedule (blue, left axis) against protein, with the posterior density of protein at the chosen nitrogen rate overlaid (orange, right axis). Most of the posterior mass sits at or past the milling step and well into the premium band, which is why the engine pays for the nitrogen: it is buying probability of clearing a step, not moving the mean protein a little."}
breaks <- attr(protein_value, "breaks")
values <- attr(protein_value, "values")
band_x <- c(min(prot_best) - 0.5, breaks, max(prot_best) + 0.5)
dens <- density(prot_best)
scale_k <- diff(range(values)) / max(dens$y)

plot_df <- rbind(
  data.frame(
    x = band_x, y = c(values, values[length(values)]),
    series = "grade-band value (AUD/t)", stringsAsFactors = FALSE
  ),
  data.frame(
    x = dens$x, y = dens$y * scale_k + min(values),
    series = "posterior density of protein", stringsAsFactors = FALSE
  )
)
ggplot2::ggplot(plot_df, ggplot2::aes(x = x, y = y, colour = series)) +
  ggplot2::geom_step(
    data = plot_df[plot_df$series == "grade-band value (AUD/t)", ],
    direction = "hv", linewidth = 1
  ) +
  ggplot2::geom_line(
    data = plot_df[plot_df$series == "posterior density of protein", ],
    linewidth = 1
  ) +
  ggplot2::geom_vline(
    xintercept = breaks, linetype = "dotted", colour = "grey60"
  ) +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.8) +
  ggplot2::scale_y_continuous(
    name = "Grade-band value (AUD/t)",
    sec.axis = ggplot2::sec_axis(
      ~ (. - min(values)) / scale_k, name = "Posterior density of protein"
    )
  ) +
  ggplot2::labs(
    x = "Protein (%)",
    title = "Grade-band schedule and posterior at chosen rate"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig-grade-band-skip, echo = FALSE, eval = !requireNamespace("ggplot2", quietly = TRUE), results = "asis"}
cat("ggplot2 is not installed; the grade-band figure is skipped.\n")
```

## Read

The nitrogen-rate decision settles on `r d_crop@action` kg N/ha, at an
expected profit of `r sprintf("%.0f", d_crop@expected_utility)` AUD/ha --
an interior point of the nine-rate grid, short of the `r max(rates)`
kg N/ha ceiling, because beyond it the marginal yield response no longer
covers the input cost. The figure makes the same point visually: the chosen
rate sits at the turning point of the expected-profit curve, not at an
extreme of it, and every rejected candidate stays visible in the ledger
above it rather than being discarded once the maximiser has run.

The firewall changes nothing about the posterior or the prices, only the
`grounding` token, and that alone is enough to move the decision:
`d_crop_ungrounded` abstains (`abstained` is
`r d_crop_ungrounded@abstained`) with reason
`` `r d_crop_ungrounded@abstain_reason` ``, returns to the status-quo rate
of `r d_crop_ungrounded@action` kg N/ha, and its action label is suffixed
`` `r grounding_unverified()` `` so the caveat travels with the number
wherever it is printed. No profit figure is quoted for a rate the evidence
does not support.

The trading decision picks a `r sprintf("%.0f", 100 * d_trade@action)`%
position (`r sprintf("$%.0f", d_trade@metadata$position)` of the
`r sprintf("$%.0f", 1000)` capital base) on a small positive edge -- exactly
the `kelly_cap` ceiling of `r sprintf("%.0f", 100 * 0.5)`%, so the binding
constraint here is the fractional-Kelly cap rather than the drawdown
guard. With no edge the same engine settles on `r flat@action`: a
zero-edge return series gives the maximiser nothing to buy. Grounding the
identical return draws as `` `r grounding_unverified()` `` reproduces the
nitrogen firewall in the trading domain: `d_trade_ungrounded` abstains
(`r d_trade_ungrounded@abstained`) for the same reason,
`` `r d_trade_ungrounded@abstain_reason` ``, forcing flat regardless of how
attractive the draws look.

The manifest tail reproduces the hand-assembled nitrogen decision exactly --
`d_tail@action` is `r d_tail@action`, matching `d_crop@action` -- because
`decide_rate_from_manifest()` reads the identical draws and the identical
grounding token, only through the manifest's public properties rather than
by hand. Its `inputs$emitter_package` records
`` `r d_tail@inputs$emitter_package` ``, the provenance carried across the
pipeline boundary. Swapping the manifest's
grounding to `` `r grounding_unverified()` `` abstains the tail
(`r d_tail_ungrounded@abstained`) exactly as the hand-assembled call did: the
firewall is enforced at the boundary of the pipeline, not only inside a
single function call.

The grade-band decision applies `r d_protein@action` kg N/ha of the
`r max(rates_n)` kg N/ha grid -- enough to be certain of clearing the
milling threshold (`P(>=11.5 milling)` is
`r sprintf("%.0f", 100 * p_milling)`%) and to clear the premium threshold on
`r sprintf("%.0f", 100 * p_premium)`% of the posterior, but not so much that
the nitrogen cost outruns the extra premium. The figure shows why: most of
the posterior density of protein at the chosen rate sits at or past the
milling step and well into the premium band, so the engine is buying
probability of clearing a step rather than nudging the mean protein a
fraction of a percentage point.

## Limits

Every posterior in this vignette is synthetic -- a diminishing-returns
nitrogen response, a Gaussian return series, a shifted protein response --
chosen to keep the vignette fast and self-contained, not fitted to a field
trial or a live price feed; a real PESTO inversion or kernR causal test
would supply the actual posterior these examples stand in for. The
grounding token is set directly by the caller in every example here, by
hand, with `grounding_grounded()` or `grounding_unverified()`; `decideR`
performs no verification step of its own. Establishing that a fact has
actually been checked against the external authority that owns it is the
job of the upstream producer, and `decideR`'s contribution is solely to
combine that provenance worst-case and abstain when the combined label is
not grounded. The trading position's `max_drawdown` guard is a hard
one-period loss-quantile constraint, not a multi-period peak-to-trough
drawdown; a decision made period by period under this guard is not
equivalent to a backtested trading strategy over a path. The `low_ess`
trigger is opt-in and is not exercised anywhere in this vignette, since no
call here supplies `min_ess`; `decisive_prob` is left at its default of
0.6 throughout, and a different threshold would move exactly which
near-tie candidates abstain as `insufficient_evidence`, a sensitivity not
explored here. The value-of-information verbs (`decide_evpi()`,
`decide_evsi()`) and the benefit-cost verbs (`decide_bcr()`,
`bcr_crrdc()`) added since this vignette was first written are not
demonstrated here; see *What to read next*. Every decision above is a
single static candidate scan, not a sequential or dynamic decision problem.

## What to read next

*From an inference result to a grain plan*, `grainPlan`'s vignette, is the
grain-specific decision orchestrator built directly on `decideR`'s manifest
tail -- it carries the pipeline demonstrated above through to a
practitioner-facing recommendation. The package's own reference pages for
`decide_evpi()`, `decide_evsi()`, `decide_bcr()` and `bcr_crrdc()` cover the
value-of-information and CRRDC-guideline benefit-cost verbs this vignette
does not exercise; `NEWS.md` records the audit findings each of them closed.

## Reproduce

Seed `1` for every random draw in this vignette, set once before the first
chunk runs. Package versions follow.

```{r session-info}
sessionInfo()
```
