---
title: "Getting started with decideR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with decideR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1L)
```

```{r}
library(decideR)
```

## What decideR does

`decideR` turns *uncertainty about the world* into *an action*. Given posterior
draws of an unknown state $\theta$, an explicit utility $U(a, \theta)$, and a set
of candidate actions, it chooses the feasible action of greatest expected utility,
$\arg\max_a \mathrm{E}_\theta[U(a, \theta)]$. It is the decision layer of the
ORCHESTRA analytics framework: the same engine sizes a crop nitrogen rate and a
trading position, because both are the same problem -- act under uncertainty, and
refuse to act when the evidence will not support it.

Two principles separate `decideR` from a bare `which.max`:

1. **Calibrated abstention.** A decision may decline to move off a safe fallback
   action when the evidence is too thin, too few, or not clearly better than the
   status quo. Abstention is a first-class outcome, recorded with a reason.
2. **The Independent Oracle Principle (IOP).** Every decision carries a
   *grounding* label, combined worst-case from its inputs. A decision built on a
   fact that has not been checked against the external authority that owns it is
   labelled `[unverified]` and is downgraded to abstention -- never asserted with
   false confidence.

## A crop decision: nitrogen rate

Suppose an upstream model -- a mechanistic-simulator inversion, say -- gives us
posterior draws of grain yield at each candidate nitrogen rate. We choose the
rate that maximises expected profit, $\mathrm{E}[p_y\,y(r) - p_r\,r]$.

```{r}
rates <- seq(0, 200, by = 25)
ymax  <- rnorm(2000L, mean = 5.5, sd = 0.3)        # posterior of asymptotic yield
yield_draws <- vapply(
  rates,
  function(r) 3 + (ymax - 3) * (1 - exp(-r / 80)) + rnorm(2000L, 0, 0.1),
  numeric(2000L)
)

d_crop <- decide_input_rate(
  yield_draws, rates, price_grain = 300, price_input = 1.2,
  grounding = grounding_grounded()
)
d_crop
```

The decision selects an interior optimum and reports the expected profit that
justified it. The full candidate ledger is available for inspection or a methods
appendix:

```{r}
head(d_crop@candidates)
```

### What happens when the yield evidence is not grounded

The same call, but now the yield draws come from a source that has *not* been
verified against an external authority (the IOP default):

```{r}
d_crop_ungrounded <- decide_input_rate(
  yield_draws, rates, price_grain = 300, price_input = 1.2,
  grounding = grounding_unverified()
)
d_crop_ungrounded
```

The decision abstains to the status quo (`rate = 0`), records the reason
`input_ungrounded`, and stamps the action label `[unverified]`. No confident
recommendation is minted on an unchecked fact.

## A trading decision: position size

The identical engine sizes a position. Given predictive draws of the next-period
return, `decide_position_size()` chooses the fraction of capital that maximises
expected log-growth $\mathrm{E}[\log(1 + f\,r)]$ -- the Kelly objective -- over a
grid capped at `kelly_cap`, subject to a hard drawdown constraint.

```{r}
returns <- rnorm(5000L, mean = 0.02, sd = 0.03)    # a small positive edge
d_trade <- decide_position_size(
  returns, capital = 1000, max_drawdown = 0.15,
  grounding = grounding_grounded()
)
d_trade
d_trade@metadata$position                          # implied dollar position
```

With **no edge**, the grid maximiser is naturally flat -- the engine does not
manufacture a position where none is warranted:

```{r}
flat <- decide_position_size(
  rnorm(5000L, mean = 0, sd = 0.03), capital = 1000, max_drawdown = 0.15,
  grounding = grounding_grounded()
)
flat@action                                        # 0: flat
```

And the **capital firewall**: an un-grounded return series -- a price feed not
verified against the exchange -- forces flat, regardless of how attractive the
numbers look.

```{r}
d_trade_ungrounded <- decide_position_size(
  returns, capital = 1000, max_drawdown = 0.15,
  grounding = grounding_unverified()
)
d_trade_ungrounded@abstained
d_trade_ungrounded@abstain_reason
```

## The shared keystone

Both decisions above route through one internal engine. A nitrogen rate and a
position size are the same computation -- expected-utility maximisation over a
candidate grid, under a feasibility constraint, with grounding and abstention.
That is the design claim `decideR` makes concrete: one decision layer, many
domains, and the same honesty discipline applied to a paddock and to real money.

## The pipeline tail: deciding from an upstream manifest

In an ORCHESTRA pipeline the yield posterior above is not assembled by hand --
it arrives as an `orchestra_manifest` emitted by an upstream member (a PESTO
ensemble inversion, a kernR causal test, a flexyBayes posterior).
`decide_rate_from_manifest()` is the standard tail of that pipeline: it reads
the predictive draws and the producer's grounding token straight off the
manifest and closes the loop into a decision. `decideR` does not depend on the
contract package -- the manifest is read through its public properties
(`outputs`, `metadata`, `summary`), so the tail consumes any conforming
producer rather than special-casing each one.

```{r}
# In production the manifest is a real `orchestra_manifest` S7 object from the
# composition layer. For a self-contained vignette we build a minimal S7
# stand-in carrying the same public properties the tail reads: the yield draws
# in `outputs` (the contract's prediction slot) and the producer's grounding
# token in `metadata$grounding`.
vignette_manifest <- S7::new_class(
  "vignette_manifest",
  properties = list(
    outputs         = S7::new_property(S7::class_any, default = NULL),
    metadata        = S7::new_property(S7::class_list, default = list()),
    summary         = S7::new_property(S7::class_any, default = NULL),
    run_id          = S7::new_property(S7::class_character, default = ""),
    emitter_package = S7::new_property(S7::class_character, default = "")
  )
)
m_grounded <- vignette_manifest(
  outputs = yield_draws,
  metadata = list(grounding = grounding_grounded()),
  run_id = "pesto-001", emitter_package = "PESTO"
)
m_ungrounded <- vignette_manifest(
  outputs = yield_draws,
  metadata = list(grounding = grounding_unverified()),
  run_id = "pesto-002", emitter_package = "PESTO"
)
```

```{r}
d_tail <- decide_rate_from_manifest(
  m_grounded, rates = rates, price_grain = 300, price_input = 1.2
)
d_tail
d_tail@inputs$emitter_package      # provenance carried from the manifest
```

The grounding rides from the producer through the decision. A manifest whose
payload is `[unverified]` forces the status quo -- the Independent Oracle
Principle firewall applied at the pipeline boundary, so `decideR` never mints a
confident rate on an upstream fact the producer did not ground.

```{r}
decide_rate_from_manifest(
  m_ungrounded, rates = rates, price_grain = 300, price_input = 1.2
)@abstained
```

For a fully general tail -- any candidate set, any utility -- use
`decide_from_manifest()`, which `decide_rate_from_manifest()` wraps.

## A threshold-economics loss: grade-band payoffs

Many grain payoffs are not smooth in the quality they reward. A delivery is
paid by *grade*: the price steps up as protein clears a milling threshold and
again at a premium threshold. The value of an outcome depends on which band it
falls in, not on its exact magnitude, so the decision is governed by the
posterior probability of clearing each band -- not by the posterior mean. A
half-percent of expected protein is worth a great deal near a band edge and
nothing in the middle of a band.

`grade_band_value()` makes that step payoff a first-class, reusable loss: a
value schedule with breakpoints.

```{r}
protein_value <- grade_band_value(
  breaks = c(11.5, 13.0),            # milling and premium protein thresholds
  values = c(300, 360, 420)          # $/t: feed, milling, premium
)
protein_value
```

Now a worked grain decision. Nitrogen lifts grain protein with diminishing
returns; the per-hectare value folds in a 5.5 t/ha yield, and nitrogen costs
1.3 per unit rate. We hold a posterior over the base (zero-nitrogen) protein and
choose the rate of greatest expected per-hectare value.

```{r}
qshift <- function(rate, base_protein) {
  base_protein + 3.0 * (1 - exp(-rate / 50))    # protein response to nitrogen
}
yield <- 5.5
grain_value <- function(rate, base_protein) {
  yield * protein_value(qshift(rate, base_protein)) - 1.3 * rate
}

base_protein <- rnorm(4000L, mean = 10.6, sd = 0.5)
rates_n <- seq(0, 150, by = 15)
d_protein <- decide(base_protein, grain_value, candidates = rates_n,
                    grounding = grounding_grounded(), safe_action = 0)
d_protein
```

The decision applies enough nitrogen to pull protein reliably over the milling
threshold -- the band premium pays for the nitrogen -- but stops short of the
rate where the nitrogen cost outruns the premium. The threshold economics is
visible in the candidate ledger: expected value rises steeply while the milling
band is being secured, then flattens.

```{r}
head(d_protein@candidates)
prot_best <- qshift(d_protein@action, base_protein)
c(`P(>=11.5 milling)` = mean(prot_best >= 11.5),
  `P(>=13 premium)`   = mean(prot_best >= 13.0))
```

For an action that shifts the quality outcome additively, `grade_band_utility()`
builds the `decide()` utility directly; supply a `quality_shift` for any
non-additive response (the saturating nitrogen response above). The grade-band
loss is the building block the future `grainPlan` orchestrator reaches for when
it prices a quality-graded decision.

## The abstention vocabulary

A decision reports exactly why it declined, if it did. The reasons, in the order
they are checked:

| Reason | Meaning |
|---|---|
| `input_ungrounded` | An input fact is `[unverified]` (the IOP firewall). |
| `no_feasible_action` | No candidate satisfies the constraint. |
| `low_ess` | The effective sample size is below the floor. |
| `insufficient_evidence` | The best action does not clearly beat the status quo. |

In every case the returned `action` is the caller's `safe_action`, and
`is_grounded()` reports whether the decision rests on grounded evidence -- the
single predicate a downstream consumer (for example a live order path) gates on.
