---
title: "Getting started with decideR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with decideR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1L)
```

```{r}
library(decideR)
```

## What decideR does

`decideR` turns *uncertainty about the world* into *an action*. Given posterior
draws of an unknown state $\theta$, an explicit utility $U(a, \theta)$, and a set
of candidate actions, it chooses the feasible action of greatest expected utility,
$\arg\max_a \mathrm{E}_\theta[U(a, \theta)]$. It is the decision layer of the
ORCHESTRA analytics framework: the same engine sizes a crop nitrogen rate and a
trading position, because both are the same problem -- act under uncertainty, and
refuse to act when the evidence will not support it.

Two principles separate `decideR` from a bare `which.max`:

1. **Calibrated abstention.** A decision may decline to move off a safe fallback
   action when the evidence is too thin, too few, or not clearly better than the
   status quo. Abstention is a first-class outcome, recorded with a reason.
2. **The Independent Oracle Principle (IOP).** Every decision carries a
   *grounding* label, combined worst-case from its inputs. A decision built on a
   fact that has not been checked against the external authority that owns it is
   labelled `[unverified]` and is downgraded to abstention -- never asserted with
   false confidence.

## A crop decision: nitrogen rate

Suppose an upstream model -- a mechanistic-simulator inversion, say -- gives us
posterior draws of grain yield at each candidate nitrogen rate. We choose the
rate that maximises expected profit, $\mathrm{E}[p_y\,y(r) - p_r\,r]$.

```{r}
rates <- seq(0, 200, by = 25)
ymax  <- rnorm(2000L, mean = 5.5, sd = 0.3)        # posterior of asymptotic yield
yield_draws <- vapply(
  rates,
  function(r) 3 + (ymax - 3) * (1 - exp(-r / 80)) + rnorm(2000L, 0, 0.1),
  numeric(2000L)
)

d_crop <- decide_input_rate(
  yield_draws, rates, price_grain = 300, price_input = 1.2,
  grounding = grounding_grounded()
)
d_crop
```

The decision selects an interior optimum and reports the expected profit that
justified it. The full candidate ledger is available for inspection or a methods
appendix:

```{r}
head(d_crop@candidates)
```

### What happens when the yield evidence is not grounded

The same call, but now the yield draws come from a source that has *not* been
verified against an external authority (the IOP default):

```{r}
d_crop_ungrounded <- decide_input_rate(
  yield_draws, rates, price_grain = 300, price_input = 1.2,
  grounding = grounding_unverified()
)
d_crop_ungrounded
```

The decision abstains to the status quo (`rate = 0`), records the reason
`input_ungrounded`, and stamps the action label `[unverified]`. No confident
recommendation is minted on an unchecked fact.

## A trading decision: position size

The identical engine sizes a position. Given predictive draws of the next-period
return, `decide_position_size()` chooses the fraction of capital that maximises
expected log-growth $\mathrm{E}[\log(1 + f\,r)]$ -- the Kelly objective -- over a
grid capped at `kelly_cap`, subject to a hard drawdown constraint.

```{r}
returns <- rnorm(5000L, mean = 0.02, sd = 0.03)    # a small positive edge
d_trade <- decide_position_size(
  returns, capital = 1000, max_drawdown = 0.15,
  grounding = grounding_grounded()
)
d_trade
d_trade@metadata$position                          # implied dollar position
```

With **no edge**, the grid maximiser is naturally flat -- the engine does not
manufacture a position where none is warranted:

```{r}
flat <- decide_position_size(
  rnorm(5000L, mean = 0, sd = 0.03), capital = 1000, max_drawdown = 0.15,
  grounding = grounding_grounded()
)
flat@action                                        # 0: flat
```

And the **capital firewall**: an un-grounded return series -- a price feed not
verified against the exchange -- forces flat, regardless of how attractive the
numbers look.

```{r}
d_trade_ungrounded <- decide_position_size(
  returns, capital = 1000, max_drawdown = 0.15,
  grounding = grounding_unverified()
)
d_trade_ungrounded@abstained
d_trade_ungrounded@abstain_reason
```

## The shared keystone

Both decisions above route through one internal engine. A nitrogen rate and a
position size are the same computation -- expected-utility maximisation over a
candidate grid, under a feasibility constraint, with grounding and abstention.
That is the design claim `decideR` makes concrete: one decision layer, many
domains, and the same honesty discipline applied to a paddock and to real money.

## The abstention vocabulary

A decision reports exactly why it declined, if it did. The reasons, in the order
they are checked:

| Reason | Meaning |
|---|---|
| `input_ungrounded` | An input fact is `[unverified]` (the IOP firewall). |
| `no_feasible_action` | No candidate satisfies the constraint. |
| `low_ess` | The effective sample size is below the floor. |
| `insufficient_evidence` | The best action does not clearly beat the status quo. |

In every case the returned `action` is the caller's `safe_action`, and
`is_grounded()` reports whether the decision rests on grounded evidence -- the
single predicate a downstream consumer (for example a live order path) gates on.
